---
title: "Mini Data Analysis (Milestone 2)"
author: "Marshall Chester"
date: "10/17/2021"
output:
  github_document:
    toc: true
    toc_depth: 2
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(datateachr)
library(tidyverse)
```

# **Milestone 2**

The focus of Milestone 2 is to address each research question through manipulating and summarizing the ```cancer_sample``` data frame.

## **Task 1**

Research questions:

1. Is ```area_mean``` a significant predictor of a malignant diagnosis?

2. Which category of ```symmetry_mean``` (low, medium, high) has the highest mean area_mean?

3. Are ```perimeter_mean```, ```smoothness_mean```, and ```symmetry_mean``` normally distributed?

4. Is ```perimeter_mean``` lognormally distributed?

### **Question 1**: Is ```area_mean``` a significant predictor of a malignant diagnosis?\n

To begin, summarized characteristics of ```area_mean``` will be tabled. 

```{r area_mean range}
tibble(cancer_sample) %>% #make new tibble to find range by category
  select(diagnosis, area_mean) %>% #select area mean
  group_by(diagnosis) %>% #group by diagnoses
  summarise(across(where(is.numeric), range)) #find range
```

```{r area_mean mean}
tibble(cancer_sample) %>% #make new tibble to find mean by category
  select(diagnosis, area_mean) %>% #select area mean
  group_by(diagnosis) %>% #group by diagnoses
  summarise(across(where(is.numeric), mean)) #find mean
```

```{r area_mean median}
tibble(cancer_sample) %>% #make new tibble to find median by category
  select(diagnosis, area_mean) %>% #select area mean
  group_by(diagnosis) %>% #group by diagnoses
  summarise(across(where(is.numeric), median)) #find median
```

```{r area_mean standard deviation}
tibble(cancer_sample) %>% #make tibble for standard deviation by category
  select(diagnosis, area_mean) %>% #select area mean
  group_by(diagnosis) %>% #group by diagnoses
  summarise(across(where(is.numeric), sd)) #find standard deviation
```

To visualize the relationship between the means of ```area_mean``` and ```diagnosis```, a boxplot will be created. 

```{r area_mean vs diagnosis boxplot, fig.align = "center"}
ggplot(cancer_sample, #visualize relationship between area and diagnosis
       aes(x = diagnosis,
           y = area_mean)) +
  geom_boxplot(aes(),
             alpha = 0.4) + 
    geom_jitter(alpha = 0.4, #customize alpha transparency
              width =0.25) + 
  labs(x = "Diagnosis", #add labels
       y = "Area (Mean)", 
       caption = "Source: UCI, 1995") +
  scale_x_discrete(breaks = c("B","M"), #change x tick
                   labels = c("Benign", "Malignant"))
```

An ANOVA test to determine the ratio of variability within and between grouped means of area_mean by diagnosis will be displayed. The summarized output will be printed.

```{r significance area_mean}
summary(aov(area_mean ~ diagnosis, data = cancer_sample)) #run ANOVA for differences of means
#f-value of 573.1 and p less than 2*10^-16
summary(lm(area_mean ~ diagnosis, data = cancer_sample)) #quantify relationship
#t-value for malignant diagnosis 23.94
```

With an F-value of 573.1 and p-value less than 0.05, there is a significant variability of group means. Furthermore, a t-value greater than 2 (t-value = 23.94, p-value < 0.05) we can say with 95% confidence that the difference in means is statistically significant. Finally, a multiple r-squared value of 0.5027 indicates that greater than 50% of the variance in the response data is explained by the model. Therefore, ```area_mean``` is a significant predictor of a malignant diagnosis. 

### **Question 2**: Which category of ```symmetry_mean``` (low, medium, high) has the highest mean area_mean?\n

To visualize the means of each category, a boxplot containing a new category of symmetry_mean will be created. 

```{r symmetry_mean category, fig.align = "center"}
cancer_sample %>% #create categorical variable
  mutate(category_symmetry = cut(symmetry_mean, 
                        breaks = 3,
                      labels = c("low","medium","high"))) %>%
  ggplot(aes(x = category_symmetry, #create boxplot
           y = area_mean)) +
  geom_boxplot(aes(),
             alpha = 0.4) +
  geom_point(aes(colour = factor(diagnosis)),
             alpha = 0.2) +
  labs(x = "Symmetry Category", #add labels
       y = "Area (Mean)", 
       caption = "Source: UCI, 1995") +
    scale_x_discrete(breaks = c("low","medium", "high"), #change x tick
                   labels = c("Low", "Medium", "High")) + 
    scale_colour_discrete(limits = c("B", "M"),
                          labels = c("Benign", "Malignant"),
                          name = "Diagnosis")
```

Therefore, it can be concluded that the category of "high" in ```symmetry_mean``` produces the highest mean ```area_mean``` across all categories.

### **Question 3**: Are ```perimeter_mean```, ```smoothness_mean```, and ```symmetry_mean``` normally distributed?\n 

Summary statistics of ```perimeter_mean``` will be displayed.

```{r summary of perimeter_mean range}
tibble(cancer_sample) %>% #make new tibble to find range by category
  select(diagnosis, perimeter_mean) %>% #select perimeter mean
  group_by(diagnosis) %>% #group by diagnoses
  summarise(across(where(is.numeric), range)) #find range
```

```{r summary of perimeter_mean mean}
tibble(cancer_sample) %>% #make new tibble to find mean by category
  select(diagnosis, perimeter_mean) %>% #select perimeter mean
  group_by(diagnosis) %>% #group by diagnoses
  summarise(across(where(is.numeric), mean)) #find mean
```

```{r summary of perimeter_mean median}
tibble(cancer_sample) %>% #make new tibble to find median by category
  select(diagnosis, perimeter_mean) %>% #select perimeter mean
  group_by(diagnosis) %>% #group by diagnoses
  summarise(across(where(is.numeric), median)) #find median
```

```{r summary of perimeter_mean standard deviation}
tibble(cancer_sample) %>% #make tibble for standard deviation by category
  select(diagnosis, perimeter_mean) %>% #select perimeter mean
  group_by(diagnosis) %>% #group by diagnoses
  summarise(across(where(is.numeric), sd)) #find standard deviation
```

To visualize the normal distribution, histograms and qq plots for each variable will be created. Different sized bins will be allocated to each histogram. 

```{r perimeter histogram, fig.align = "center"}
ggplot(cancer_sample,
       aes(perimeter_mean)) +
  geom_histogram(bins = 30) + 
    labs(x = "Perimeter (Mean)", #add labels
       y = "Count", 
       caption = "Source: UCI, 1995")
```

```{r perimeter qqplot, fig.align = "center"}
ggplot(cancer_sample,
       aes(sample = perimeter_mean)) +
  stat_qq() + stat_qq_line()
```

```{r smoothness histogram, fig.align = "center"}
ggplot(cancer_sample,
       aes(smoothness_mean)) +
  geom_histogram(bins = 25) + 
    labs(x = "Smoothness (Mean)", #add labels
       y = "Count", 
       caption = "Source: UCI, 1995")
```

```{r smoothness qqplot, fig.align = "center"}
ggplot(cancer_sample,
       aes(sample = smoothness_mean)) +
  stat_qq() + stat_qq_line()
```

```{r symmetry histogram, fig.align = "center"}
ggplot(cancer_sample,
       aes(symmetry_mean)) +
  geom_histogram(bins = 20) + 
    labs(x = "Symmetry (Mean)", #add labels
       y = "Count", 
       caption = "Source: UCI, 1995")
```

```{r symmetry qqplot, fig.align = "center"}
ggplot(cancer_sample,
       aes(sample = symmetry_mean)) +
  stat_qq() + stat_qq_line()
```

The ```perimeter_mean``` histogram appears to be lognormal with concentrated values below the mean and right skew. The qq plot appears to show non-normal deviation from the qq line. The ```smoothness_mean``` and ```symmetry_mean``` histograms appear normal with concentrated values about the mean and no skew. The qq plots show normal distribution.

### **Question 4**: Is ```perimeter_mean``` lognormally distributed?

The number of observations in the malignant variable will be calculated.

```{r malignant observations}
tibble(cancer_sample) %>% #make tibble
  select(diagnosis, perimeter_mean) %>% 
  filter(diagnosis == "M") %>% #filter by diagnoses
  count(diagnosis) # count number of observations
```

```{r perimeter log-transformed histogram, fig.align = "center"}
ggplot(cancer_sample, #make logarithmic graph
       aes(perimeter_mean,
           log = "y")) +
  geom_histogram(bins = 25) + 
  scale_y_continuous(trans = "log10") + 
    labs(x = "Perimeter (Mean)", #add labels
       y = "Count", 
       caption = "Source: UCI, 1995")
```

```{r symmetry log-transformed qqplot, fig.align = "center"}
ggplot(cancer_sample,
       aes(sample = symmetry_mean),
         scale_y_continuous(trans = "log10")) +
  stat_qq() + stat_qq_line()
```

After log transformation, the ```perimeter_mean``` histogram appears to be concentrated about the mean without skew. The log-transformed qq plot now follows a distribution more indicative of a normal distribution. Therefore, the ```perimeter_mean``` variable is log-normally distributed. 

### **Conclusions**

The research questions have been slightly altered from Milestone 1 to produce reasonably conclusive findings for Milestone 2. At this juncture, ```area_mean``` and ```symmetry_mean``` are producing the most interesting results and a further investigation into the relationship between these variables and ```diagnosis``` is most compelling to the author. To initiate this investigation, the research questions will be refined in a later section of the report.

## **Task 2**

### **Tidy Data**

All datasets should be tidy and reshaped to ensure: 

* each row is an observation
* each column is a variable
* each cell is a value

The following tibble used for the analysis of each research question is in tidy format. 

```{r cancer tibble subset original tidy, fig.align = "centre"}
tibble(cancer_sample) %>% 
  select(diagnosis, radius_mean, perimeter_mean, area_mean, smoothness_mean, symmetry_mean) %>%
  mutate(category_symmetry = cut(symmetry_mean, 
                        breaks = 3,
                      labels = c("low","medium","high")))
```

In the ```cancer_sample``` dataset, each row is an observation. Each column is a unique variable of the tumour, and each cell is a value in the form of numbers (quantitative) or strings (qualitative). In its current format, the data meets the criteria of tidy data. 

In this next example, the data will be untidied with the addition of categorical and continuous variables in the ```radius_mean``` column.

```{r cancer tibble subset untidy, fig.align = "centre"}
tibble(cancer_sample) %>% 
  select(diagnosis, radius_mean, perimeter_mean, area_mean, smoothness_mean, symmetry_mean) %>%
  mutate(category_symmetry = cut(symmetry_mean, 
                        breaks = 3,
                        labels = c("low","medium","high"))) %>%
  mutate(radius_mean = sample(c("male","female", 18:85), #add untidy column 
                              size = 569,
                              rep = TRUE))
```

The data is will now be tidied back to original format. 

```{r cancer tibble subset tidy, fig.align = "centre"}
tibble(cancer_sample) %>% 
  select(diagnosis, radius_mean, perimeter_mean, area_mean, smoothness_mean, symmetry_mean) %>%
  mutate(radius_mean = cancer_sample$radius_mean) #return to original data set
```
### **Select final questions**

The questions proposed in Milestone 2 are essentially for the purpose of completing Milestone 3, questions 1 and 2 will are selectd and will be refined for further analysis. For question 1, additional variables (```symmetry_mean```) will be added to create predictive models. For question 2, an analysis of the significance of the difference in means can be postulated. Model building and tests of significance between means are of the greatest interest to the author. 

**Proposed additions to questions**:

1. Are ```area_mean``` and ```symmetry_mean``` both significant predictors of a malignant diagnosis?\n

```{r cancer subset question 1 tibble, fig.align = "centre"}
tibble(cancer_sample) %>%
  mutate(-radius_mean) %>% #remove column
  select(diagnosis, area_mean, symmetry_mean) %>% #select only these variables
  filter(area_mean < 2300) %>% #eliminate potential outliers
  arrange(desc(area_mean))
```
2. Is the difference of means between the categories of ```symmetry_mean``` (low, medium, high) statistically significant?\n

```{r cancer subset question 2 tibble, fig.align = "centre"}
tibble(cancer_sample) %>%
   mutate(category_symmetry = cut(symmetry_mean, 
                        breaks = 3,
                      labels = c("low","medium","high"))) %>%
  select(diagnosis, symmetry_mean, category_symmetry) %>% # select only these variables
  filter(symmetry_mean < 0.3) %>% #eliminate potential outliers
  arrange(desc(symmetry_mean))
```