---
title: "Mini Data Analysis (Milestone 1)"
author: "Marshall Chester"
date: "09/10/2021"
output:
  html_document:
    toc: true
    toc_float: true
  github_document:
    toc: true
    toc_depth: 2
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(datateachr)
library(tidyverse)
```

# **Introduction**

### **Purpose**

The purpose of this markdown is to complete STAT 545A Mini-Data Analysis (Milestone 1).\n

The author will:

1. Become familiar with one dataset
2. Create four questions from selected dataset
3. Write a reproducible and clear report with R Markdown

Several open-sourced datasets will be explored, and one final dataset will be visualized for research purposes. 

### **Author**

The author of this code is Marshall Chester, with open sourced data from the ```datateachr``` library.

### **Packages**

The following packages must be installed:

```{r packages}
# install.packages("devtools")
# devtools::install_github("UBC-MDS/datateachr")
```

The following libraries must be loaded:

```{r libraries}
library(datateachr)
library(tidyverse)
```

# **Milestone 1**

## **Task 1: Choose dataset**


### **Initial ranking**

1. flow_sample
2. cancer_sample
3. vancouver_trees
4. apt_buildings

### **Dataset exploration**

```{r exploration}
dim(flow_sample)
class(flow_sample)

dim(cancer_sample)
class(cancer_sample)

dim(vancouver_trees)
class(vancouver_trees)

dim(apt_buildings)
class(apt_buildings)

```

### **Dataset attributes**

Dataset | Rows | Columns | Class | Potential Research Question 
:------:|:----:|:-------:|:-----:|:-----------------|
flow_sample | 218 | 7 | data.frame | Is maximum flow always within June and July?
cancer_sample | 569 | 32 | data.frame | What is the average area of malignant diagnoses?
vancouver_trees | 146,611 | 20 | data.frame | Which Vancouver neighbourhood has the greatest number of tree species?
apt_buildings | 3,455 | 37 | data.frame | What percentage of buildings built before 2000 have air conditioning?

### **Pair down**

After viewing the data, I decided to proceed with cancer_sample and apt_buildings. These datasets are rich and include greater than 30 variables and 500 points for exploration and visualization.

### **Final decision**

I have decided to proceed with the cancer_sample dataset due to personal interest in healthcare. The dataset is robust with a significant number of variables and points for further evaluation. With subsetting, I believe this dataset is well suited for analysis of strength of association and predictive modeling.

## **Task 2: Explore dataset**

In the cancer_sample dataset, area_mean and radius_mean are continuous variables and can be explored by plotting each variable on the x and y axis, respectively. A positive, linear relationship is expected. By grouping data points with diagnosis (dichotomous variable), an interesting relationship could be seen.

### **New tibble**

The cancer_sample includes 32 unique variables. A new tibble can be created to include the variables most interesting to the author. For the purpose of model building, this parsed tibble will help to create an efficient workflow. 

```{r cancer tibble subset, fig.align = "centre"}
tibble(cancer_sample) %>% #make new tibble of interested variables
  select(diagnosis, radius_mean, perimeter_mean, area_mean, smoothness_mean, symmetry_mean)
```

### **Density plot**

Peaks within a density help to understand where the data are concentrated. In the cancer_sample dataset, the area_mean variable provides continuous data on the average area of diagnoses. A density plot can be created to visualize the concentration of values for both malignant and benign diagnoses. With this visualization, a greater understanding is available. 

```{r density plot, fig.align = "center"}
ggplot(cancer_sample, #check density plot of area_mean
       aes(area_mean, fill = diagnosis)) +
  geom_density(alpha = 0.6) +
  labs(x = "Area (Mean)", #add labels
       y = "Density",
       fill = "Diagnosis",
       caption = "Source: UCI, 1995") +
  scale_fill_discrete( #change legend titles
    limits = c("B", "M"),
                      labels = c("Benign", "Malignant"))
```

### **Create boxplot**

A boxplot can provide detailed information between dichotomous and continuous variables. In the cancer_sample dataset, diagnosis is dichotomous (malignant or benign) and symmetry_mean is continuous. A boxplot will help visualize the relationship between malignant and benign diagnoses. There may be an interesting outcome for further consideration.

```{r boxplot, fig.align = "center"}
ggplot(cancer_sample, #create boxplot between 2 variables
       aes(x = diagnosis,
           y = symmetry_mean)) +
  geom_boxplot(alpha = 0, #add transparency and change width
               width = 0.25) +
  geom_jitter(alpha = 0.3, #add transparency and change width
              width =0.25) + 
  labs(x = "Diagnosis", #add labels
       y = "Symmetry (Mean)", 
       caption = "Source: UCI, 1995") +
  scale_x_discrete(breaks = c("B","M"), #change x tick
                   labels = c("Benign", "Malignant"))
```

### **Explore relationship**

```{r scatterplot, fig.align = "center"}
ggplot(cancer_sample, #explore relationship between variables
       aes(x = area_mean,
           y = radius_mean, 
           group = diagnosis)) +
  geom_point(aes(color = diagnosis),
             size = 3, #add transparency and change size
             alpha = 0.4) +
    labs(x = "Area (Mean)", #add labels
         y = "Radius (Mean)", 
         caption = "Source: UCI, 1995") +
    scale_color_discrete(name = "Diagnosis", #change legend
                         labels=c("Benign", "Malignant")) +
    facet_wrap(~diagnosis) + #facet
  theme(strip.text.x = element_blank())
```

## **Task 3: Research Questions**

With the dataset selected and explored, four potential research questions are postulated.

1. Is area_mean a significant predictor of a malignant diagnosis?\n

2. Which category of symmetry_mean (low, medium, high) has the highest mean area_mean when grouped by diagnosis?\n

3. Are perimeter_mean, smoothness_mean, and symmetry_mean normally distributed?\n

4. Is perimeter_mean lognormally distributed?\n

# **Milestone 2**

Addressing each research question through manipulating and summarizing the cancer_sample data frame.  

## **Task 1**

**Question 1**: Is area_mean a significant predictor of a malignant diagnosis?\n

To visualize the relationship of area_mean and malignant diagnoses, the range, mean, median, and standard deviation across both outcomes will be computed.

```{r malignant tibbles}
tibble(cancer_sample) %>% #make new tibble to find range by category
  select(diagnosis, area_mean) %>% #filter by malignant diagnoses
  group_by(diagnosis) %>%
  summarise(across(where(is.numeric), range)) # find range

tibble(cancer_sample) %>% #make new tibble to find mean by category
  select(diagnosis, area_mean) %>% #filter by malignant diagnoses
  group_by(diagnosis) %>%
  summarise(across(where(is.numeric), mean)) # find mean

tibble(cancer_sample) %>% #make new tibble to find median by category
  select(diagnosis, area_mean) %>% #filter by malignant diagnoses
  group_by(diagnosis) %>%
  summarise(across(where(is.numeric), median)) # find median

tibble(cancer_sample) %>% #make tibble for standard deviation by category
  select(diagnosis, area_mean) %>% #filter by malignant diagnoses
  group_by(diagnosis) %>%
  summarise(across(where(is.numeric), sd)) # find standard deviation
```

```{r area_mean vs diagnosis boxplot}
ggplot(cancer_sample, #visualize relationship between area and diagnosis
       aes(x = diagnosis,
           y = area_mean)) +
  geom_boxplot(aes(),
             alpha = 0.4) + 
    geom_jitter(alpha = 0.4, #customize alpha transparency
              width =0.25) + 
  labs(x = "Diagnosis", #add labels
       y = "Area (Mean)", 
       caption = "Source: UCI, 1995") +
  scale_x_discrete(breaks = c("B","M"), #change x tick
                   labels = c("Benign", "Malignant"))
summary(aov(area_mean ~ diagnosis, data = cancer_sample)) #run ANOVA for differences of means
#f-value of 573.1 and p less than 2*10^-16
summary(lm(area_mean ~ diagnosis, data = cancer_sample)) #quantify relationship
#t-value for malignant diagnosis 23.94
```

**Question 2**: Which category of symmetry_mean (low, medium, high) has the highest mean area_mean?\n

```{r symmetry_mean category}
cancer_sample %>% 
  mutate(category_symmetry = cut(symmetry_mean, 
                        breaks = 3,
                      labels = c("low","medium","high"))) %>%
  ggplot(aes(x = category_symmetry,
           y = area_mean)) +
  geom_boxplot(aes(),
             alpha = 0.4) +
  geom_point(aes(colour = factor(diagnosis)),
             alpha = 0.2) +
  labs(x = "Symmetry Category", #add labels
       y = "Area (Mean)", 
       caption = "Source: UCI, 1995") +
    scale_x_discrete(breaks = c("low","medium", "high"), #change x tick
                   labels = c("Low", "Medium", "High")) + 
    scale_colour_discrete(limits = c("B", "M"),
                          labels = c("Benign", "Malignant"),
                          name = "Diagnosis")
```

**Question 3**: Are perimeter_mean, smoothness_mean, and symmetry_mean normally distributed?\n

```{r perimeter histogram, fig.align = "center"}
ggplot(cancer_sample,
       aes(perimeter_mean)) +
  geom_histogram(bins = 25) + 
    labs(x = "Perimeter (Mean)", #add labels
       y = "Count", 
       caption = "Source: UCI, 1995")
```

```{r perimeter qqplot, fig.align = "center"}
ggplot(cancer_sample,
       aes(sample = perimeter_mean)) +
  stat_qq() + stat_qq_line()
```

```{r smoothness histogram, fig.align = "center"}
ggplot(cancer_sample,
       aes(smoothness_mean)) +
  geom_histogram(bins = 25) + 
    labs(x = "Smoothness (Mean)", #add labels
       y = "Count", 
       caption = "Source: UCI, 1995")
```

```{r smoothness qqplot, fig.align = "center"}
ggplot(cancer_sample,
       aes(sample = smoothness_mean)) +
  stat_qq() + stat_qq_line()
```

```{r symmetry histogram, fig.align = "center"}
ggplot(cancer_sample,
       aes(symmetry_mean)) +
  geom_histogram(bins = 25) + 
    labs(x = "Symmetry (Mean)", #add labels
       y = "Count", 
       caption = "Source: UCI, 1995")
```

```{r symmetry qqplot, fig.align = "center"}
ggplot(cancer_sample,
       aes(sample = symmetry_mean)) +
  stat_qq() + stat_qq_line()
```

4. Is perimeter_mean lognormally distributed?\n

```{r perimeter log-transformed histogram, fig.align = "center"}
ggplot(cancer_sample,
       aes(perimeter_mean,
           log = "y")) +
  geom_histogram(bins = 25) + 
  scale_y_continuous(trans = "log10") + 
    labs(x = "Perimeter (Mean)", #add labels
       y = "Count", 
       caption = "Source: UCI, 1995")
```

```{r symmetry log-transformed qqplot, fig.align = "center"}
ggplot(cancer_sample,
       aes(sample = symmetry_mean),
         scale_y_continuous(trans = "log10")) +
  stat_qq() + stat_qq_line()
```

## **Task 2**

