---
title: "mda_mchester"
author: "mchester"
date: "05/10/2021"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scrool: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(datateachr)
library(tidyverse)
```

# **Introduction**

### **Purpose**

The purpose of this code is to complete STAT 545A mini-data analysis, Milestone 1. 

### **Author**

The author of this code is Marshall Chester, with open sourced data.

### **Packages**

The following packages must be installed:

- [x] install.packages("devtools")
- [x] devtools::install_github("UBC-MDS/datateachr")

The following libraries must be loaded:

- [x] library(datateachr)
- [x] library(tidyverse)


# **Task 1: Choose dataset**


### **First dataset ranking**

1. flow_sample
2. cancer_sample
3. vancouver_trees
4. apt_buildings

### **Dataset exploration**

```{r}
dim(flow_sample)
class(flow_sample)

dim(cancer_sample)
class(cancer_sample)

dim(vancouver_trees)
class(vancouver_trees)

dim(apt_buildings)
class(apt_buildings)

```

### **Dataset attributes**

Dataset | Rows | Columns | Class | Potential Research Question 
:------:|:----:|:-------:|:-----:|:-----------------|
flow_sample | 218 | 7 | data.frame | Is maximum flow always within June and July?
cancer_sample | 569 | 32 | data.frame | What is the average area of a malignant tumour in UCI population?
vancouver_trees | 146,611 | 20 | data.frame | Which Vancouver neighbourhood has the greatest number of tree species?
apt_buildings | 3,455 | 37 | data.frame | What percentage of buildings built before 2000 have air conditioning?

### **Pair down**

After viewing the data, I decided to proceed with cancer_sample and apt_buildings. The research questions formulated from the exploration were of greatest interest.

### **Final decision**

I have decided to proceed with the cancer_sample dataset due to personal interest in the research question and robust nature of the data.

# **Task 2: Explore dataset**

### **Explore relationship**
```{r, fig.align = "center"}
ggplot(cancer_sample, 
       aes(area_mean, radius_mean)) +
  geom_point(aes(color = diagnosis),
             size = 3, 
             alpha = 0.4)
```

### **Create boxplot**

```{r, fig.align = "center"}
ggplot(cancer_sample, 
       aes(x = diagnosis, y = symmetry_mean)) +
  geom_boxplot(alpha = 0) +
  geom_jitter(alpha = 0.3, colour = "tomato")
```

### **New tibble**

```{r, fig.align = "centre"}
tibble(cancer_sample) %>%
  select(diagnosis, radius_mean, area_mean, smoothness_mean, symmetry_mean)
```

### **Distribution plot**
```{r, fig.align = "center"}
ggplot(cancer_sample,
       aes(area_mean)) +
  geom_density(aes(fill = diagnosis))
```

# **Task 3: Research Questions**