---
title: "mda_mchester"
author: "mchester"
date: "05/10/2021"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(datateachr)
library(tidyverse)
```

# **Introduction**

### **Purpose**

The purpose of this markdown is to complete STAT 545A mini-data analysis, Milestone 1. Multiple open-sourced datasets will be explored, and one final dataset will be selected.

### **Author**

The author of this code is Marshall Chester, with open sourced data from the datateachr library.

### **Packages**

The following packages must be installed:

```r
install.packages("devtools")
devtools::install_github("UBC-MDS/datateachr")
```

The following libraries must be loaded:

```r
library(datateachr)
library(tidyverse)
```

# **Task 1: Choose dataset**


### **First dataset ranking**

1. flow_sample
2. cancer_sample
3. vancouver_trees
4. apt_buildings

### **Dataset exploration**

```{r}
dim(flow_sample)
class(flow_sample)

dim(cancer_sample)
class(cancer_sample)

dim(vancouver_trees)
class(vancouver_trees)

dim(apt_buildings)
class(apt_buildings)

```

### **Dataset attributes**

Dataset | Rows | Columns | Class | Potential Research Question 
:------:|:----:|:-------:|:-----:|:-----------------|
flow_sample | 218 | 7 | data.frame | Is maximum flow always within June and July?
cancer_sample | 569 | 32 | data.frame | What is the average area of malignant diagnoses?
vancouver_trees | 146,611 | 20 | data.frame | Which Vancouver neighbourhood has the greatest number of tree species?
apt_buildings | 3,455 | 37 | data.frame | What percentage of buildings built before 2000 have air conditioning?

### **Pair down**

After viewing the data, I decided to proceed with cancer_sample and apt_buildings. The research questions formulated from the exploration were of greatest interest.

### **Final decision**

I have decided to proceed with the cancer_sample dataset due to personal interest in the research question, as well as the robust and deep nature of the data and its variables.

# **Task 2: Explore dataset**

In the cancer_sample dataset, area_mean and radius_mean are continuous variables and can be explored by plotting each variable on the x and y axis, respectively. A positive, linear relationship is expected. By grouping data points with diagnosis (a dichotomous variable), an interesting plot relationship could form. 

### **Explore relationship**
```{r, fig.align = "center"}
ggplot(cancer_sample, #explore relationship between variables
       aes(x = area_mean,
           y = radius_mean, 
           group = diagnosis)) +
  geom_point(aes(color = diagnosis),
             size = 3, #add transparency and change size
             alpha = 0.4) +
    labs(x = "Area (Mean)", #add labels
         y = "Radius (Mean)", 
         caption = "Source: UCI, 1995") +
    scale_color_discrete(name = "Diagnosis", #change legend
                         labels=c("Benign", "Malignant")) +
    facet_wrap(~diagnosis) + #facet by row
  theme(strip.text.x = element_blank())
```

### **Create boxplot**

A boxplot can provide detailed information between dichotomous and continuous variables. In the cancer_sample dataset, diagnosis is dichotomous (malignant or benign) and symmetry_mean is continuous. A boxplot will help visualize the relationship between malignant and benign diagnoses. There may be an interesting outcome for further consideration.

```{r, fig.align = "center"}
ggplot(cancer_sample, #create boxplot between 2 variables
       aes(x = diagnosis,
           y = symmetry_mean)) +
  geom_boxplot(alpha = 0, #add transparency and change width
               width = 0.25) +
  geom_jitter(alpha = 0.3, #add transparency and change width
              width =0.25) + 
  labs(x = "Diagnosis", #add labels
       y = "Symmetry (Mean)", 
       caption = "Source: UCI, 1995") +
  scale_x_discrete(breaks = c("B","M"), #change x tick
                   labels = c("Benign", "Malignant"))
```

### **New tibble**

The cancer_sample includes 32 unique variables. A new tibble can be created to include the variables most interesting to the author. For the purpose of model building, this parsed tibble will help to create an efficient workflow. 

```{r, fig.align = "centre"}
tibble(cancer_sample) %>% #make new tibble of interesting variables
  select(diagnosis, radius_mean, perimeter_mean, area_mean, smoothness_mean, symmetry_mean)
```

### **Density ridge plot**

Peaks within a density help to understand where the data are concentrated. In the cancer_sample dataset, the area_mean variable provides continuous data on the average area of diagnoses. A density plot can be created to visualize the concentration of values for both malignant and benign diagnoses. With this visualization, a greater understanding is available. 

```{r, fig.align = "center"}
ggplot(cancer_sample, #check density plot of area_mean
       aes(area_mean, fill = diagnosis)) +
  geom_density(alpha = 0.6) +
  labs(x = "Area (Mean)", #add labels
       y = "Density",
       fill = "Diagnosis",
       caption = "Source: UCI, 1995") +
  scale_fill_discrete( #change legend titles
    limits = c("B", "M"),
                      labels = c("Benign", "Malignant"))
```

# **Task 3: Research Questions**

With the dataset selected and explored, four potential research questions can be formed. 

1. Which subsetted variables (see tibble) are most predictive of a malignant diagnosis?\n

2. Which subsetted variables (see tibble) are most predictive of a benign diagnosis?\n

3. Is the relationship between symmetry and malignant diagnosis statistically significant?\n

4. What is the strength of correlation between mean radius and area?\n